Draft_Application_Stochastic_Volatility

### 2021-07-19 Real Dataset SV.Rmd
commit

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Bisognerebbe fare riferimento alla sezione sulla convergenza. Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare già di convergenza quando N è molto grande) } Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target. 
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such a comparison: trivially, the latter assumes that the competing algorithms are approximating the same target distribution.

@@

As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse. With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.

>>>>

Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.
Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Bisognerebbe fare riferimento alla sezione sulla convergenza. Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare già di convergenza quando N è molto grande) } Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target. 
With such a benchmark, we can then compare how different filters approach it as the number of particles increases. We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark. Note that the Liu and West filter can not be included in such a comparison: trivially, the latter assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we also conduct this analysis only on one of the three indices, the SP500. Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}

@@

As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse. With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.
As it turns out, such suggestion is confirmed by running the algorithms with N=1000. Passing from 100 to 1000 particles nearly halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a sizable hedge over the Auxiliary PF. Such a hedge almost vanishes when increasing the number of particles to 5000: according to the RMSE, the Bootstrap PF performs slightly better than the Guided PF, which in turn yields very similar results to the Auxiliary PF.
We would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others.
